## **Data model Project**

## **Predict survival of patients with heart failure**

Ref:<https://bmcmedinformdecismak.biomedcentral.com/articles/10.1186/s12911-020-1023-5>

Cardiovascular diseases (CVDs) are disorders of the heart and blood vessels including, heart attacks, strokes, heart failure

Cardiovascular diseases cause the death of approximately 17 million people worldwide annually. In particular, heart failure occurs when the heart is unable to pump enough blood to the body, and it is usually caused by diabetes, high blood pressure, or other heart conditions or diseases.

Objective: FIND PEOPLE WITH A HIGH PROBABILITY OF DEATH TO PROVIDE IMMEDIATE ASSISTANCE.

Dataset: <https://www.kaggle.com/andrewmvd/heart-failure-clinical-data>

\
![](https://lh4.googleusercontent.com/pPPqtuGWYROd18n8JytqgmuhW7OKL5rjjzETq0UFWtb0PnJPWrLJo3Dlu1RONYGgj0d7Js5N7FRvGFuvF-qxQn0nwysZy6exadAzTySO-uHU5KTkPnCllewDqAcMfrqWoqhhB1R-Knk)

![](https://lh3.googleusercontent.com/W0kQrVvjZV8e6cjZUKhvri20WbesjP8Pl3NTSFWBzrCKumve-heVowuAPau46EEx-Fijp47kvk7DBPB2663bfTTpEu4VJ7qdxsTVzSTTcYCQTWVUa5xJKQgJ5MJzG0j_dwpcqbbnffw)

This dataset containing the medical records of 299 heart failure patients collected at the Faisalabad Institute of Cardiology and at the Allied Hospital in Faisalabad (Punjab, Pakistan), during April--December 2015

1.  **Age :** Age of the patient (Year)

2.  **Anaemia :** Decrease of red blood cells or hemoglobin (Yes = 1, No = 0)

3.  **High blood pressure :** If a patient has hypertension (Yes = 1, No = 0)

4.  **Creatinine phosphokinase :** Level of the CPK enzyme in the blood (mcg/L)

5.  **Diabetes :** If the patient has diabetes (Yes = 1, No = 0)

6.  **Ejection fraction :** Percentage of blood leaving the heart at each contraction (Percentage)

7.  **Sex :** Woman or man (Man = 1, Woman=0)

8.  **Platelets :** Platelets in the blood (kilo platelets/mL)

9.  **Serum creatinine :** Level of creatinine in the blood (mg/dL)

10. **Serum sodium :** Level of sodium in the blood (mEq/L)

11. **Smoking :** If the patient smokes (Yes = 1, No = 0)

12. **Time :** Follow-up period (Days)

#### **Death event :**If the patient died during the follow-up period (Yes = 1, No = 0)

------------------------------------------------------------------------

#### import package

```{r}
library(tidyverse)
library(caret)
library(lift)
library(ROCR)
library(rpart)
library(rpart.plot)
library(readr)
set.seed(69)
```

```{r}
heart_failure_clinical_records_dataset <- read_csv("heart_failure_dataset.csv")
heart_failure_clinical_records_dataset
```

## Data Preparation

1.Clean up data by removing Na data.

```{r}
any(is.na(heart_failure_clinical_records_dataset))
```

2.Rename column "creatinine_phosphokinase" to "cpk" and "DEATH_EVENT" to "death" and [reject time((Follow-up period (Days)) isn't used to create a model because it is an unpredictable future story.]{.ul} .

```{r}
heart_failure_clinical_records_dataset %>% rename(
    cpk = creatinine_phosphokinase,
    death = DEATH_EVENT
  ) %>% 
  select(-time) -> all_data

```

3.Change datatype of "death" to factor

```{r}
all_data$death = as.factor(all_data$death)
```

Now. dataset is ready

```{r}
all_data
```

## Data Exploration

data type

```{r}
unique(all_data$age)        #Continuous
unique(all_data$anaemia)    #Categorical
unique(all_data$cpk)        #Continuous
unique(all_data$diabetes)   #Categorical
unique(all_data$ejection_fraction)   #Continuous
unique(all_data$high_blood_pressure) #Categorical
unique(all_data$platelets)           #Continuous 
unique(all_data$serum_creatinine)    #Continuous 
unique(all_data$serum_sodium)        #Continuous
unique(all_data$sex)     #Categorical
unique(all_data$smoking) #Categorical 
unique(all_data$death)   #Response y is Categorical
```

|     Categorical     |        Numerical         |
|:-------------------:|:------------------------:|
|       Anaemia       |           Age            |
| High blood pressure | Creatinine phosphokinase |
|      diabetes       |    Ejection fraction     |
|         sex         |        platelets         |
|       smoking       |     Serum creatinine     |
|                     |       Serum sodium       |

: Type of data

```{r}
summary(all_data$death)
```

A total of **299** patients were surveyed: **203** Alive and **96** Died

```{r}
ggplot(all_data, aes(age)) + 
  geom_histogram(binwidth = 5,colour = "black", fill="salmon2")
```

## Model explanation

#### 1.Decision tree Hold out method 70% train 30%

#### Train

```{r}
#Separate train and test data
n <- nrow(all_data)
test_index <- sample(1:n,size =  0.3*n) 
data_train <- all_data[-test_index,] #train 0.7
data_test <- all_data[test_index,]   #test 0.3
```

```{r}
#Summary train and test dataset
summary(data_test$death)
summary(data_train$death)
```

```{r}
#Train model
model1 <- rpart(death~., data = data_train)
```

```{r}
rpart.plot(model1)
```

```{r}
model1$variable.importance
```

```{r}
summary(model1)$cptable
```

## **Test**

```{r}
#Test Model
res1 <- predict(model1, data_test, type = 'class') 
#class type return factor ('0', '1')
```

```{r}
head(res1)
confusionMatrix(res1, data_test$death, positive = "1")
confusionMatrix(res1, data_test$death, positive = "1", mode = 'prec_recall')
```

#### Evaluation : Lift Calculation of Model1

```{r}
#Lift Calculation of Model1
res1_pos <- predict(model1, data_test)[,'1']
lift_result1 <- data.frame(prob = res1_pos, y = data_test$death)
lift_obj1 <- lift(y ~ prob, data = lift_result1, class = '1')
plot(lift_obj1) #add parameter values (determine %sample found) to find %sample tested
```

```{r}
pred1 <- prediction(res1_pos, data_test$death, label.ordering = c('0', '1'))
perf_lift1 <- performance(pred1, 'lift', 'rpp')
plot(perf_lift1)

tdl1 <- TopDecileLift(res1_pos, as.integer(data_test$death)-1) 
tdl1
```

#### 

#### 2.Cross Validation (CV)

```{r}
#Cross Validation
train_control <- trainControl(method = 'cv', number = 10)
model_cv <- train(death~. ,
                  data = data_train, 
                  trControl = train_control, 
                  method = 'rpart')
model_cv
model_cv$finalModel #get new model
```

```{r}
model_final <- model_cv$finalModel
rpart.plot(model_final)
model_final$variable.importance
```

```{r}
#Test Model from CV
res2 <- predict(model_final, data_test, type = 'class')
head(res2)
confusionMatrix(res2, data_test$death, positive = "1")
confusionMatrix(res2, data_test$death, positive = "1", mode = 'prec_recall')
```

```{r}
#Lift Calculation of Model from CV
res2_pos <- predict(model_final, data_test)[,'1']
lift_result2 <- data.frame(prob = res2_pos, y = data_test$death)
lift_obj2 <- lift(y ~ prob, data = lift_result2, class = '1')
plot(lift_obj2, values = 60) #add parameter values (determine %sample found) to find %sample tested
```

```{r}
pred2 <- prediction(res2_pos, data_test$death, label.ordering = c('0', '1'))
perf_lift2 <- performance(pred2, 'lift', 'rpp')
plot(perf_lift2) 

tdl2 <- TopDecileLift(res2_pos, as.integer(data_test$death)-1) 
tdl2
```

#### 3.Cross Validation (LOOCV)

```{r}
#Cross Validation LOOCV
train_control_one <- trainControl(method = 'LOOCV', number = 10)
model_cv_one <- train(death~. ,
                  data = all_data, 
                  trControl = train_control_one, 
                  method = 'rpart')
```

```{r}
model_cv_one
model_cv_one$finalModel #get new model
```

```{r}
model_final_one <- model_cv_one$finalModel
rpart.plot(model_final_one)
model_final_one$variable.importance
```

```{r}
#Test Model from LOOCV
res2_one <- predict(model_final_one, data_test, type = 'class')
head(res2_one)
confusionMatrix(res2_one, data_test$death, positive = "1")
confusionMatrix(res2_one, data_test$death, positive = "1", mode = 'prec_recall')
```

```{r}
#Lift Calculation of Model from CV
res3_pos <- predict(model_final_one, data_test)[,'1']
lift_result3 <- data.frame(prob = res3_pos, y = data_test$death)
lift_ob32 <- lift(y ~ prob, data = lift_result3, class = '1')
plot(lift_obj2, values = 60) #add parameter values (determine %sample found) to find %sample tested
```

```{r}
pred3 <- prediction(res3_pos, data_test$death, label.ordering = c('0', '1'))
perf_lift3 <- performance(pred3, 'lift', 'rpp')
plot(perf_lift3) 

tdl3 <- TopDecileLift(res3_pos, as.integer(data_test$death)-1) 
tdl3
```
